---
title: "Constructing country-round inequality exposure"
subtitle: "Interview-year weighted SWIID and OECD inequality measures for ESS"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Purpose

This notebook documents the construction of country-round inequality exposure measures for the *Inequality and social cohesion* project.

The European Social Survey (ESS) is organised in survey rounds, but fieldwork for a given country-round can span more than one calendar year. Because the contextual inequality measures are annual, assigning a single calendar-year value to an ESS country-round can misrepresent the inequality context actually experienced by respondents.

The approach used here therefore:

1. starts from the observed distribution of ESS interviews across calendar years within each country-round;
2. calculates the share of respondents interviewed in each year;
3. uses those shares to form weighted country-round inequality measures;
4. treats SWIID as the provisional primary inequality source;
5. retains OECD Income Distribution Database (IDD) measures for robustness analyses; and
6. preserves the full ESS country-round structure even when contextual inequality data are unavailable.

The resulting dataset contains one row per ESS country-round and is saved as:

`03_data/interim/ess_country_round_inequality.csv`

## Setup

```{r}
library(dplyr)
library(readr)
library(here)
```

## 1. Import the ESS country-round structure

Two previously constructed project files are used.

The first contains the full ESS country-round structure:

```{r}
country_rounds <- read_csv(
  here(
    "03_data",
    "interim",
    "ess_country_round_coverage.csv"
  ),
  show_col_types = FALSE
)
```

The second contains one row per country × ESS round × interview year, already matched to the annual OECD and SWIID inequality series:

```{r}
coverage <- read_csv(
  here(
    "03_data",
    "interim",
    "ess_inequality_coverage.csv"
  ),
  show_col_types = FALSE
)
```

A compact overview of the inputs:

```{r}
tibble(
  object = c(
    "country_rounds",
    "coverage"
  ),
  n_rows = c(
    nrow(country_rounds),
    nrow(coverage)
  ),
  n_columns = c(
    ncol(country_rounds),
    ncol(coverage)
  )
)
```

The distinction between these two inputs matters:

- `country_rounds` defines the **complete ESS country-round universe** used by the project.
- `coverage` contains the subset of country-round-year observations for which interview-year timing is available and annual inequality measures can be attached.

This allows missing contextual coverage to remain visible rather than causing ESS country-rounds to disappear silently.

## 2. Why interview-year weighting is necessary

Some ESS country-rounds span two calendar years. Suppose, for example, that 80% of interviews in a country-round were completed in 2014 and 20% in 2015. Using only the round's nominal year or a simple average of 2014 and 2015 would give equal influence to both years even though respondents were not interviewed equally across them.

Instead, each interview year receives a weight equal to its share of respondents in the relevant country-round:

$$
w_{cty} =
\frac{N_{cty}}
{\sum_y N_{cty}}
$$

where:

- $c$ denotes country,
- $t$ denotes ESS round,
- $y$ denotes interview year, and
- $N_{cty}$ is the number of respondents interviewed in year \(y\).

The country-round contextual value is then:

$$
X_{ct} = \sum_y w_{cty}X_{cy}
$$

This makes the contextual measure correspond to the observed timing of the ESS interviews.

## 3. Calculate interview-year shares

```{r}
coverage <- coverage |>
  group_by(cntry, essround) |>
  mutate(
    interview_year_share =
      n_respondents / sum(n_respondents)
  ) |>
  ungroup()
```

### Verify that the weights sum to one

```{r}
share_check <- coverage |>
  group_by(cntry, essround) |>
  summarise(
    share_sum = sum(interview_year_share),
    .groups = "drop"
  )
```

```{r}
share_check |>
  summarise(
    min_share_sum = min(share_sum),
    max_share_sum = max(share_sum),
    max_absolute_deviation =
      max(abs(share_sum - 1))
  )
```

The construction requires each country-round's interview-year shares to sum to one. The following assertion stops execution if that condition is violated:

```{r}
stopifnot(
  all(abs(share_check$share_sum - 1) < 1e-10)
)
```

## 4. Aggregate annual inequality to the ESS country-round level

The aggregation creates one row per country-round.

Several pieces of timing information are retained alongside the inequality measures so that the temporal construction remains transparent.

```{r}
inequality_country_round <- coverage |>
  group_by(cntry, essround) |>
  summarise(

    # Fieldwork timing
    n_respondents_with_year =
      sum(n_respondents),

    first_interview_year =
      min(interview_year),

    last_interview_year =
      max(interview_year),

    mean_interview_year =
      weighted.mean(
        interview_year,
        interview_year_share
      ),

    # SWIID: primary series
    gini_swiid_contemp =
      weighted.mean(
        gini_swiid_contemp,
        interview_year_share
      ),

    gini_swiid_lag1 =
      weighted.mean(
        gini_swiid_lag1,
        interview_year_share
      ),

    # Average reported SWIID uncertainty.
    # Diagnostic only; NOT used as an inferential SE.
    swiid_lag1_se_mean =
      weighted.mean(
        gini_swiid_lag1_se,
        interview_year_share
      ),

    # OECD: retain a country-round value only when every
    # interview-year component has observed OECD data.
    oecd_contemp_complete =
      all(!is.na(gini_oecd_contemp)),

    oecd_lag1_complete =
      all(!is.na(gini_oecd_lag1)),

    gini_oecd_contemp =
      if (all(!is.na(gini_oecd_contemp))) {
        weighted.mean(
          gini_oecd_contemp,
          interview_year_share
        )
      } else {
        NA_real_
      },

    gini_oecd_lag1 =
      if (all(!is.na(gini_oecd_lag1))) {
        weighted.mean(
          gini_oecd_lag1,
          interview_year_share
        )
      } else {
        NA_real_
      },

    .groups = "drop"
  )
```

## 5. Interpretation of the constructed variables

### Fieldwork timing

The following variables document how each country-round was formed:

- `n_respondents_with_year`: number of respondents contributing to the interview-year weighting;
- `first_interview_year`: earliest observed interview year;
- `last_interview_year`: latest observed interview year;
- `mean_interview_year`: interview-share weighted mean calendar year.

`mean_interview_year` is useful as a compact description of the effective timing of a country-round, but it is not itself used to look up contextual data.

### SWIID measures

Two SWIID measures are constructed:

- `gini_swiid_contemp`: contemporaneous inequality;
- `gini_swiid_lag1`: one-year lagged inequality.

SWIID is treated as the provisional primary inequality source because it provides substantially broader country-year coverage than OECD IDD.

The notebook also retains:

- `swiid_lag1_se_mean`

This is the interview-share weighted mean of the SWIID-reported uncertainty measure attached to the lagged Gini values.

It is retained **for diagnostics only**. It is not an inferential standard error for the country-round estimate and should not be used as though it were one. Propagating SWIID uncertainty requires a separate multiple-imputation-style analysis using the SWIID draws.

### OECD measures

OECD IDD is retained for robustness analyses.

A stricter completeness rule is used for OECD aggregation. If a country-round spans multiple interview years, an OECD country-round value is calculated only when **all** contributing interview years have an observed OECD value.

For example, if a country-round has interviews in 2014 and 2015 but OECD data are available only for 2014, the country-round OECD measure is set to missing rather than representing the whole round with the available year alone.

The indicators:

- `oecd_contemp_complete`
- `oecd_lag1_complete`

make this rule explicit.

## 6. Inspect the aggregated country-round data

```{r}
inequality_country_round |>
  arrange(cntry, essround) |>
  slice_head(n = 20)
```

A quick summary of coverage before restoring the full ESS structure:

```{r}
inequality_country_round |>
  summarise(
    n_country_rounds = n(),
    n_swiid_contemp =
      sum(!is.na(gini_swiid_contemp)),
    n_swiid_lag1 =
      sum(!is.na(gini_swiid_lag1)),
    n_oecd_contemp =
      sum(!is.na(gini_oecd_contemp)),
    n_oecd_lag1 =
      sum(!is.na(gini_oecd_lag1))
  )
```

## 7. Restore the full ESS country-round structure

The aggregation above begins from the interview-year coverage file. Any ESS country-round with no usable interview-year information would therefore be absent unless the complete country-round structure were restored explicitly.

At the time this workflow was constructed, **Estonia Round 5** had no usable interview-year information and was absent from `coverage`.

The country-round inequality data are therefore joined back onto the full ESS country-round table:

```{r}
country_round_inequality <- country_rounds |>
  select(
    cntry,
    essround,
    n_respondents,
    first_year,
    last_year,
    n_interview_years,
    crosses_year
  ) |>
  left_join(
    inequality_country_round,
    by = c("cntry", "essround")
  ) |>
  arrange(cntry, essround)
```

This design choice ensures that missing contextual information remains an explicit missing value rather than silently changing the ESS country-round sample.

## 8. Diagnostics

### Overall country-round coverage

```{r}
coverage_summary <- tibble(
  measure = c(
    "ESS country-rounds",
    "Country-rounds with SWIID lagged Gini",
    "Country-rounds with complete OECD lagged Gini"
  ),
  n = c(
    nrow(country_round_inequality),
    sum(!is.na(
      country_round_inequality$gini_swiid_lag1
    )),
    sum(!is.na(
      country_round_inequality$gini_oecd_lag1
    ))
  )
)

coverage_summary
```

### Missing SWIID country-rounds

```{r}
missing_swiid <- country_round_inequality |>
  filter(is.na(gini_swiid_lag1)) |>
  select(
    cntry,
    essround,
    n_respondents,
    first_year,
    last_year
  )

missing_swiid
```

The explicit display of missing country-rounds is an important diagnostic because the main analysis should not silently lose observations through contextual-data merges.

## 9. Compare SWIID and OECD where both are observed

OECD IDD is not used as the primary source, but overlapping country-rounds provide a useful robustness check on whether the two sources describe broadly similar cross-national inequality patterns.

```{r}
source_comparison <- country_round_inequality |>
  filter(
    !is.na(gini_swiid_lag1),
    !is.na(gini_oecd_lag1)
  )
```

```{r}
source_correlation <- cor(
  source_comparison$gini_swiid_lag1,
  source_comparison$gini_oecd_lag1
)

tibble(
  n_overlapping_country_rounds =
    nrow(source_comparison),
  correlation_swiid_oecd_lag1 =
    source_correlation
)
```

This correlation is descriptive. Agreement between sources does not imply that they are interchangeable: SWIID and OECD differ in construction, country coverage, and treatment of missing annual observations. OECD is therefore retained as a robustness source rather than averaged together with SWIID.

## 10. Save the country-round inequality dataset

```{r}
write_csv(
  country_round_inequality,
  here(
    "03_data",
    "interim",
    "ess_country_round_inequality.csv"
  )
)
```

The saved file contains one row per ESS country-round and is the contextual inequality input used by later analysis-data construction scripts.

## Summary of methodological decisions

This construction encodes several substantive and reproducibility decisions:

1. Contextual inequality is aligned to actual ESS interview timing. Country-round values are respondent-share weighted across observed interview years rather than assigned from a single nominal round year.

2. SWIID is the provisional primary source. Its broader temporal and cross-national coverage makes it suitable for the main analysis.

3. Both contemporaneous and one-year lagged inequality are retained. The lagged measure is intended to reduce temporal simultaneity between the contextual exposure and respondent outcomes.

4. SWIID uncertainty is retained only as a diagnostic in this file. The weighted mean reported uncertainty is not treated as an inferential standard error. Proper propagation of SWIID uncertainty requires modelling across the underlying SWIID draws.

5. OECD IDD is a robustness source. Country-round OECD estimates are retained only when every interview-year component has observed OECD data.

6. Incomplete OECD country-rounds are not partially averaged. This prevents a multi-year ESS round from being represented by only the subset of interview years for which OECD data happen to be available.

7. The complete ESS country-round structure is restored after aggregation. Missing contextual data remain explicit rather than changing the sample silently.

8. Source agreement is inspected but not used to combine sources. SWIID and OECD remain analytically distinct.

## Next steps

The resulting `ess_country_round_inequality.csv` is intended to be merged into the respondent-level ESS analysis data by `cntry` and `essround`.

Later stages of the project can then compare:

- contemporaneous versus lagged inequality;
- SWIID versus OECD specifications; and
- point-estimate analyses versus analyses that propagate SWIID uncertainty across its underlying imputations.
