---
title: "Constructing post-Round-9 contextual exposures"
subtitle: "Interview-timing-weighted inequality and macroeconomic context for ESS Rounds 10–11"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-copy: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Purpose and status

This notebook documents the construction of country-round contextual exposures for European Social Survey (ESS) Rounds 10 and 11.

The authoritative executable script is:

```text
04_code/01_import-clean/07_construct_post9_context.R
```

The construction extends the contextual-data logic developed for Rounds 1–9 to the post-R9 ESS period while preserving actual interview timing.

The contextual variables are:

- SWIID disposable-income Gini;
- World Bank GDP per capita, PPP;
- World Bank unemployment.

For each measure, both contemporaneous and one-year-lagged country-round values are constructed.

The primary specification uses the one-year lag.

This step does **not** append Rounds 10 and 11 respondents to the analysis dataset. That occurs subsequently in:

```text
04_code/01_import-clean/08_build_ess_extended_analysis.R
```

## Inputs

The construction uses:

```text
03_data/interim/ess_post9_country_round_year_coverage.csv
03_data/interim/ess_post9_country_round_coverage.csv
03_data/interim/ess_inequality_coverage.csv

03_data/raw/inequality/swiid/
03_data/raw/macro/world_bank/
```

The first two files describe actual ESS interview timing in Rounds 10 and 11.

The Rounds 1–9 inequality-coverage file is used to reuse the established ESS-to-SWIID country correspondence.

## Outputs

The main country-round output is:

```text
03_data/interim/ess_post9_context.csv
```

Additional diagnostics are written to:

```text
03_data/interim/ess_post9_context_year_coverage.csv
03_data/interim/ess_post9_context_annual_coverage_summary.csv
03_data/interim/ess_post9_context_country_round_coverage_summary.csv
03_data/interim/ess_post9_context_complete_summary.csv
03_data/interim/ess_post9_context_gaps.csv
```

## Run the authoritative construction

Rather than maintain a second copy of the full construction code in this notebook, the notebook executes the authoritative R script and then exposes the resulting objects and diagnostics.

```{r}
#| label: run-post9-context
#| include: false

library(dplyr)
library(here)
library(readr)
library(tidyr)

source(
  here(
    "04_code",
    "01_import-clean",
    "07_construct_post9_context.R"
  )
)
```

## 1. Actual ESS interview timing

ESS country-rounds are not assigned a single nominal survey year.

Instead, the contextual exposure uses the observed distribution of interviews across calendar years.

For country $c$, ESS round $t$, and interview year $y$, define:

$$
s_{cty}
=
\frac{n_{cty}}
{\sum_y n_{cty}},
$$

where $n_{cty}$ is the number of respondents with a usable interview year in that country-round-year cell.

The shares therefore sum to one among respondents with known interview timing.

```{r}
#| label: post9-year-shares

year_shares |>
  select(
    cntry,
    essround,
    interview_year,
    n_respondents,
    n_known_date,
    n_total,
    interview_year_share,
    known_date_share,
    lag_year
  ) |>
  arrange(
    essround,
    cntry,
    interview_year
  )
```

The construction also retains `known_date_share`, the proportion of respondents in the country-round with usable interview-year information.

This separates two issues:

1. how interviews with known dates are distributed across years;
2. how completely interview timing is observed in the country-round.

```{r}
#| label: incomplete-interview-timing

year_shares |>
  distinct(
    cntry,
    essround,
    n_total,
    n_known_date,
    known_date_share
  ) |>
  filter(
    known_date_share < 1
  ) |>
  arrange(
    known_date_share,
    essround,
    cntry
  )
```

Respondents without a usable interview year are not assigned to an assumed calendar year.

The contextual value is based on the observed timing distribution among respondents for whom timing is known.

## 2. One-year-lagged contextual exposure

For a contextual variable $X$, the one-year-lagged country-round exposure is:

$$
X^{lag1}_{ct}
=
\sum_y
s_{cty}
X_{c,y-1}.
$$

The contemporaneous exposure is:

$$
X^{current}_{ct}
=
\sum_y
s_{cty}
X_{cy}.
$$

This means that a country-round spanning several interview years can incorporate several annual contextual observations.

For example, if 80% of dated respondents were interviewed in 2021 and 20% in 2022, the lagged Gini exposure is:

$$
0.80 \times Gini_{2020}
+
0.20 \times Gini_{2021}.
$$

## 3. SWIID inequality

The primary inequality source is SWIID version 9.92.

The construction uses the disposable-income Gini coefficient.

ESS country codes are mapped explicitly to SWIID country names. The established Rounds 1–9 mapping is reused and supplemented where required for countries appearing only in the later ESS rounds.

The construction uses `swiid_summary` for these summary-series contextual exposures.

An important scale distinction is retained:

- `gini_disp` in `swiid_summary` is already expressed on the conventional 0–100 Gini scale;
- `gini_disp` in the individual SWIID imputation objects is stored in hundredths and must be divided by 100.

The latter distinction becomes important in the separate SWIID uncertainty workflow.

## 4. Macroeconomic controls

GDP and unemployment are drawn from World Bank country-year data.

The current indicators are:

| Construct | World Bank indicator | Transformation |
|---|---|---|
| GDP per capita, PPP | `NY.GDP.PCAP.PP.KD` | Natural logarithm |
| Unemployment | `SL.UEM.TOTL.ZS` | None |

GDP per capita is measured in constant 2021 international dollars.

The unemployment series is the modelled ILO estimate as a percentage of the total labour force.

ESS two-letter country codes are matched to World Bank ISO3 codes through an explicit project crosswalk.

Both contemporaneous and one-year-lagged annual values are attached before country-round aggregation.

## 5. Annual contextual-data coverage

Coverage is assessed before aggregation.

```{r}
#| label: annual-context-coverage

annual_coverage_summary
```

This distinction matters because a country-round may span more than one interview year.

Having one annual contextual value available is not sufficient if another required annual component is missing.

## 6. Complete-component rule

Country-round contextual exposures are constructed only when the contextual source provides values for **every** interview-year component required by the observed fieldwork timing.

The operative rule in the R script is equivalent to:

```r
weighted_if_complete <- function(x, w) {
  if (all(!is.na(x))) {
    sum(x * w)
  } else {
    NA_real_
  }
}
```

This is a consequential design choice.

The workflow does not:

- interpolate missing annual source values;
- extrapolate beyond the end of a source series;
- drop missing years and renormalise the remaining interview-year weights.

Suppose a country-round has:

$$
s_{ct,2022}=0.2
$$

and

$$
s_{ct,2023}=0.8.
$$

If the lagged contextual source contains the required 2021 value but not the required 2022 value, the country-round receives a missing lagged exposure.

It is **not** treated as if all respondents had been exposed to the 2021 value.

This prevents source-data availability from changing the substantive timing represented by the country-round exposure.

## 7. Country-round exposures

The constructed post-R9 context contains:

- `gini_swiid_current`;
- `gini_swiid_lag1`;
- `log_gdp_pc_ppp_current`;
- `log_gdp_pc_ppp_lag1`;
- `unemployment_current`;
- `unemployment_lag1`.

The primary contextual-data completeness flag is:

```r
complete_primary_context =
  !is.na(gini_swiid_lag1) &
  !is.na(log_gdp_pc_ppp_lag1) &
  !is.na(unemployment_lag1)
```

This flag corresponds to the contextual variables required by the macro-adjusted models.

```{r}
#| label: country-round-context-coverage

country_round_coverage_summary
```

```{r}
#| label: complete-context-summary

complete_context_summary
```

## 8. Contextual-data gaps

The construction explicitly retains country-rounds whose contextual values cannot be completed.

```{r}
#| label: post9-context-gaps

context_gaps |>
  select(
    cntry,
    essround,
    n_respondents,
    known_date_share,
    gini_swiid_lag1,
    log_gdp_pc_ppp_lag1,
    unemployment_lag1,
    missing_gini,
    missing_gdp,
    missing_unemployment
  ) |>
  arrange(
    essround,
    cntry
  )
```

These rows should not be interpreted as processing failures.

The subsequent source audit traces the missingness to identifiable source-series or timing limitations.

The relevant audit script is:

```text
04_code/02_descriptives/08_context_source_audit.R
```

For the complete Rounds 1–11 macro-context sample, the later-round exclusions arise primarily because the one-year lag requires source values beyond the available endpoint of the local SWIID or macroeconomic series.

No missing annual components are filled by extrapolation.

## 9. Relationship to the extended ESS dataset

The output `ess_post9_context.csv` is subsequently read by:

```text
04_code/01_import-clean/08_build_ess_extended_analysis.R
```

That script:

1. harmonises ESS Rounds 10 and 11 to the Rounds 1–9 respondent structure;
2. attaches the post-R9 lagged SWIID exposure;
3. combines baseline and later-round macro context;
4. creates the Rounds 1–11 respondent-level analysis dataset;
5. retains interview mode for sensitivity analysis.

The principal extended respondent-level output is:

```text
03_data/processed/ess_analysis_r1_r11.rds
```

The corresponding combined contextual files are:

```text
03_data/interim/ess_country_round_macro_r1_r11.csv
03_data/interim/ess_country_round_context_r1_r11.csv
```

## 10. Interpretation

The post-R9 construction preserves the same substantive timing principle used in the baseline analysis:

> contextual conditions are attached according to when ESS respondents were actually interviewed, rather than according to a nominal survey-round year.

The one-year-lag specification therefore represents a timing-weighted approximation to contextual conditions preceding the observed survey interviews.

It should not be interpreted as a precise individual-level exposure date.

The country-round remains the contextual unit.

## 11. Reproducibility notes

The notebook depends on locally available ESS, SWIID, and World Bank source data that are not distributed through the public repository.

The executable R script remains authoritative for the construction itself.

This notebook documents the methodological choices that are especially important for interpretation:

- use of actual interview timing;
- explicit country-source mappings;
- one-year lagging;
- respondent-share aggregation;
- the complete-component rule;
- preservation of source-data gaps rather than interpolation or renormalisation.