---
title: "Building the extended ESS analysis dataset"
subtitle: "Harmonising ESS Rounds 10–11 with the Rounds 1–9 baseline"
date: today
date-format: "08.09.2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Purpose

This notebook documents the construction of the extended European Social Survey (ESS) analysis dataset covering **Rounds 1–11**.

The project began with a cleaned respondent-level dataset covering ESS Rounds 1–9. Later rounds were added separately because the post-R9 period differs both substantively and operationally from the earlier ESS series.

The extension is therefore designed to:

1. preserve the existing `ess_analysis.rds` file as the R1–R9 baseline;
2. harmonise interviewer-administered Rounds 10 and 11 to the same analytical structure;
3. attach one-year-lagged SWIID inequality exposure to the later rounds;
4. retain macroeconomic controls in a separate country-round file;
5. preserve interview mode for later sensitivity analyses; and
6. verify that respondent counts, identifiers, and contextual joins remain intact throughout the extension.

The main outputs are:

- `03_data/processed/ess_analysis_r1_r11.rds`
- `03_data/interim/ess_country_round_macro_r1_r11.csv`
- `03_data/interim/ess_country_round_context_r1_r11.csv`
- respondent- and context-level diagnostic CSV files

## Setup

```{r}
library(dplyr)
library(haven)
library(here)
library(readr)
```

## 1. Helper functions

Three small helper functions are used repeatedly when harmonising the later ESS rounds.

### Numeric extraction

`get_num()` returns a numeric version of a variable when present and otherwise returns a vector of missing values of the appropriate length.

This makes the harmonisation robust to variables that are absent from one of the source files.

```{r}
get_num <- function(
  dat,
  variable
) {

  if (
    variable %in%
      names(dat)
  ) {

    as.numeric(
      zap_labels(
        dat[[variable]]
      )
    )

  } else {

    rep(
      NA_real_,
      nrow(dat)
    )
  }
}
```

### Character extraction

`get_chr()` performs the analogous operation for character variables.

```{r}
get_chr <- function(
  dat,
  variable
) {

  if (
    variable %in%
      names(dat)
  ) {

    as.character(
      dat[[variable]]
    )

  } else {

    rep(
      NA_character_,
      nrow(dat)
    )
  }
}
```

### Range cleaning

`clean_range()` retains only values inside an expected valid range.

This is used for variables such as trust, age, and education to prevent special codes or unexpected numeric values from entering the harmonised dataset.

```{r}
clean_range <- function(
  x,
  lower,
  upper
) {

  if_else(
    !is.na(x) &
      x >= lower &
      x <= upper,
    x,
    NA_real_
  )
}
```

## 2. Import the existing R1–R9 analysis dataset

The previously cleaned Rounds 1–9 dataset remains the baseline.

```{r}
ess_base_raw <- readRDS(
  here(
    "03_data",
    "processed",
    "ess_analysis.rds"
  )
)
```

```{r}
tibble(
  dataset = "Existing R1-R9 analysis data",
  n_respondents = nrow(ess_base_raw),
  n_variables = ncol(ess_base_raw)
)
```

### Verify required baseline variables

The extension assumes that several core analytical variables already exist in the baseline file.

```{r}
required_base_vars <- c(
  "cntry",
  "essround",
  "idno",
  "ppltrst",
  "age",
  "gender",
  "education_years",
  "analysis_weight",
  "gini_swiid_lag1"
)

missing_base_vars <- setdiff(
  required_base_vars,
  names(ess_base_raw)
)

if (
  length(missing_base_vars) > 0
) {

  stop(
    paste(
      "Required variables missing from ess_analysis.rds:",
      paste(
        missing_base_vars,
        collapse = ", "
      )
    )
  )
}
```

This check prevents the extension from running against an incompatible version of the baseline dataset.

## 3. Standardise the R1–R9 structure

The baseline file is reduced to the structure that will also be produced for Rounds 10 and 11.

This standardisation is important because `bind_rows()` should combine conceptually equivalent variables rather than merely columns with similar names.

```{r}
ess_base <- tibble(
  essround =
    as.integer(
      ess_base_raw$essround
    ),

  cntry =
    as.character(
      ess_base_raw$cntry
    ),

  idno =
    as.numeric(
      ess_base_raw$idno
    ),

  interview_year =
    if (
      "interview_year" %in%
        names(ess_base_raw)
    ) {
      as.numeric(
        ess_base_raw$interview_year
      )
    } else {
      NA_real_
    },

  interview_month =
    if (
      "interview_month" %in%
        names(ess_base_raw)
    ) {
      as.numeric(
        ess_base_raw$interview_month
      )
    } else {
      NA_real_
    },

  # Rounds 1-9 use the conventional face-to-face design.
  interview_mode =
    "In-person",

  interview_mode_raw =
    "Face-to-face",

  ppltrst =
    as.numeric(
      ess_base_raw$ppltrst
    ),

  pplfair =
    get_num(
      ess_base_raw,
      "pplfair"
    ),

  pplhlp =
    get_num(
      ess_base_raw,
      "pplhlp"
    ),

  age =
    as.numeric(
      ess_base_raw$age
    ),

  gender =
    as.character(
      ess_base_raw$gender
    ),

  education_years =
    as.numeric(
      ess_base_raw$education_years
    ),

  # Labour-force variables retained but not yet harmonised
  pdwrk =
    get_num(
      ess_base_raw,
      "pdwrk"
    ),

  edctn =
    get_num(
      ess_base_raw,
      "edctn"
    ),

  uempla =
    get_num(
      ess_base_raw,
      "uempla"
    ),

  uempli =
    get_num(
      ess_base_raw,
      "uempli"
    ),

  dsbld =
    get_num(
      ess_base_raw,
      "dsbld"
    ),

  rtrd =
    get_num(
      ess_base_raw,
      "rtrd"
    ),

  cmsrv =
    get_num(
      ess_base_raw,
      "cmsrv"
    ),

  hswrk =
    get_num(
      ess_base_raw,
      "hswrk"
    ),

  dngoth =
    get_num(
      ess_base_raw,
      "dngoth"
    ),

  mainact =
    get_num(
      ess_base_raw,
      "mainact"
    ),

  mnactic =
    get_num(
      ess_base_raw,
      "mnactic"
    ),

  # Household economic position
  hincfel =
    get_num(
      ess_base_raw,
      "hincfel"
    ),

  hinctnta =
    get_num(
      ess_base_raw,
      "hinctnta"
    ),

  # Migration/background
  brncntr =
    get_num(
      ess_base_raw,
      "brncntr"
    ),

  ctzcntr =
    get_num(
      ess_base_raw,
      "ctzcntr"
    ),

  analysis_weight =
    as.numeric(
      ess_base_raw$analysis_weight
    ),

  gini_swiid_lag1 =
    as.numeric(
      ess_base_raw$gini_swiid_lag1
    )
)
```

The labour-force variables are deliberately retained without further harmonisation here. Their presence keeps open the possibility of using them later, but this script does not yet impose a new labour-market classification across all rounds.

## 4. Import Rounds 10 and 11

The later rounds are stored as separate interviewer-administered ESS extracts.

```{r}
r10_file <- here(
  "03_data",
  "raw",
  "ess",
  "post9",
  "ess10_f2f.dta"
)

r11_file <- here(
  "03_data",
  "raw",
  "ess",
  "post9",
  "ess11_f2f.dta"
)

stopifnot(
  file.exists(r10_file),
  file.exists(r11_file)
)
```

```{r}
ess10_raw <- read_dta(
  r10_file
)

ess11_raw <- read_dta(
  r11_file
)
```

```{r}
tibble(
  essround = c(10, 11),
  n_respondents = c(
    nrow(ess10_raw),
    nrow(ess11_raw)
  ),
  n_variables = c(
    ncol(ess10_raw),
    ncol(ess11_raw)
  )
)
```

Keeping the raw later-round files separate from the original R1–R9 pipeline makes the post-R9 extension transparent and reversible.

## 5. Harmonise the later-round respondent data

The `clean_post9()` function converts a Round 10 or 11 source file into the same respondent-level structure used for the baseline data.

### Why a dedicated harmonisation function is used

The later rounds require several explicit decisions:

- extract interview year and month from the ESS interview-date variable;
- preserve and standardise interview mode;
- clean trust and demographic variables to valid ranges;
- harmonise gender labels;
- reconstruct analysis weights when `ANWEIGHT` is unavailable;
- retain additional labour, household, and migration variables where present; and
- protect against missing or unexpected `essround` coding.

```{r}
clean_post9 <- function(
  dat,
  round_number
) {

  # Interview date
  inwds <-
    dat[["inwds"]]

  if (
    !inherits(
      inwds,
      c(
        "POSIXct",
        "POSIXt",
        "Date"
      )
    )
  ) {

    stop(
      paste(
        "inwds is not a recognised date/time variable in Round",
        round_number
      )
    )
  }

  # Interview-mode labels
  mode_raw <- if (
    "mode" %in%
      names(dat)
  ) {

    as.character(
      as_factor(
        dat$mode
      )
    )

  } else {

    rep(
      NA_character_,
      nrow(dat)
    )
  }

  mode_standard <- case_when(
    grepl(
      "video",
      mode_raw,
      ignore.case = TRUE
    ) ~
      "Video",

    grepl(
      "face|capi|in.person|personal",
      mode_raw,
      ignore.case = TRUE
    ) ~
      "In-person",

    is.na(mode_raw) ~
      NA_character_,

    TRUE ~
      mode_raw
  )

  # Core identifiers
  essround <-
    get_num(
      dat,
      "essround"
    )

  cntry <-
    get_chr(
      dat,
      "cntry"
    )

  idno <-
    get_num(
      dat,
      "idno"
    )

  # Social-trust outcomes
  ppltrst <-
    clean_range(
      get_num(
        dat,
        "ppltrst"
      ),
      0,
      10
    )

  pplfair <-
    clean_range(
      get_num(
        dat,
        "pplfair"
      ),
      0,
      10
    )

  pplhlp <-
    clean_range(
      get_num(
        dat,
        "pplhlp"
      ),
      0,
      10
    )

  # Age
  age <-
    clean_range(
      get_num(
        dat,
        "agea"
      ),
      15,
      120
    )

  # Gender
  gndr <-
    get_num(
      dat,
      "gndr"
    )

  gender <- case_when(
    gndr == 1 ~
      "Man",

    gndr == 2 ~
      "Woman",

    TRUE ~
      NA_character_
  )

  # Education
  education_years <-
    clean_range(
      get_num(
        dat,
        "eduyrs"
      ),
      0,
      50
    )

  # Analysis weights:
  # use ANWEIGHT when available; otherwise reconstruct
  # from PSPWGHT × PWEIGHT.
  anweight <-
    get_num(
      dat,
      "anweight"
    )

  pspwght <-
    get_num(
      dat,
      "pspwght"
    )

  pweight <-
    get_num(
      dat,
      "pweight"
    )

  fallback_weight <-
    pspwght *
    pweight

  analysis_weight <- case_when(
    !is.na(anweight) &
      anweight > 0 ~
      anweight,

    !is.na(fallback_weight) &
      fallback_weight > 0 ~
      fallback_weight,

    TRUE ~
      NA_real_
  )

  tibble(
    essround =
      as.integer(essround),

    cntry,

    idno,

    interview_year =
      as.numeric(
        format(
          inwds,
          "%Y"
        )
      ),

    interview_month =
      as.numeric(
        format(
          inwds,
          "%m"
        )
      ),

    interview_mode =
      mode_standard,

    interview_mode_raw =
      mode_raw,

    ppltrst,
    pplfair,
    pplhlp,

    age,
    gender,
    education_years,

    # Labour/activity
    pdwrk =
      get_num(
        dat,
        "pdwrk"
      ),

    edctn =
      get_num(
        dat,
        "edctn"
      ),

    uempla =
      get_num(
        dat,
        "uempla"
      ),

    uempli =
      get_num(
        dat,
        "uempli"
      ),

    dsbld =
      get_num(
        dat,
        "dsbld"
      ),

    rtrd =
      get_num(
        dat,
        "rtrd"
      ),

    cmsrv =
      get_num(
        dat,
        "cmsrv"
      ),

    hswrk =
      get_num(
        dat,
        "hswrk"
      ),

    dngoth =
      get_num(
        dat,
        "dngoth"
      ),

    mainact =
      get_num(
        dat,
        "mainact"
      ),

    mnactic =
      get_num(
        dat,
        "mnactic"
      ),

    # Household economic position
    hincfel =
      get_num(
        dat,
        "hincfel"
      ),

    hinctnta =
      get_num(
        dat,
        "hinctnta"
      ),

    # Migration/background
    brncntr =
      get_num(
        dat,
        "brncntr"
      ),

    ctzcntr =
      get_num(
        dat,
        "ctzcntr"
      ),

    analysis_weight
  ) |>
    mutate(
      # Protect against unexpected ESSROUND coding.
      essround =
        if_else(
          is.na(essround),
          as.integer(round_number),
          essround
        ),

      interview_year =
        if_else(
          interview_year >= 2020 &
            interview_year <= 2025,
          interview_year,
          NA_real_
        ),

      interview_month =
        if_else(
          interview_month >= 1 &
            interview_month <= 12,
          interview_month,
          NA_real_
        )
    )
}
```

## 6. Apply the harmonisation to Rounds 10 and 11

```{r}
ess10 <- clean_post9(
  ess10_raw,
  10
)

ess11 <- clean_post9(
  ess11_raw,
  11
)

ess_post9 <- bind_rows(
  ess10,
  ess11
)
```

A compact check:

```{r}
ess_post9 |>
  count(essround, name = "n_respondents")
```

## 7. Attach post-R9 inequality exposure

The one-year-lagged SWIID Gini for Rounds 10 and 11 was previously constructed in:

`03_data/interim/ess_post9_context.csv`

That file also contains later-round macroeconomic context.

```{r}
post9_context <- read_csv(
  here(
    "03_data",
    "interim",
    "ess_post9_context.csv"
  ),
  show_col_types = FALSE
)
```

### Verify contextual uniqueness before joining

A country-round contextual file must contain at most one row per `cntry × essround`.

```{r}
context_duplicates <- post9_context |>
  count(
    cntry,
    essround
  ) |>
  filter(
    n > 1
  )

if (
  nrow(context_duplicates) > 0
) {

  stop(
    "Duplicate country-rounds found in ess_post9_context.csv."
  )
}
```

### Join lagged inequality to respondents

```{r}
ess_post9 <- ess_post9 |>
  left_join(
    post9_context |>
      select(
        cntry,
        essround,
        gini_swiid_lag1
      ),
    by = c(
      "cntry",
      "essround"
    )
  )
```

A left join is used because the respondent-level ESS sample is the backbone. Missing contextual values should remain visible rather than causing respondents to be dropped.

## 8. Combine R1–R9 with R10–R11

```{r}
ess_extended <- bind_rows(
  ess_base,
  ess_post9
) |>
  mutate(
    post_r9 =
      essround >= 10,

    period =
      case_when(
        essround <= 9 ~
          "R1-R9",

        essround == 10 ~
          "R10",

        essround == 11 ~
          "R11",

        TRUE ~
          NA_character_
      )
  ) |>
  arrange(
    essround,
    cntry,
    idno
  )
```

The `post_r9` and `period` variables are retained explicitly because later models may compare the original R1–R9 period with R10 and R11 rather than assuming a single homogeneous ESS series.

## 9. Integrity checks

The extended dataset should preserve every respondent in both the baseline and later-round inputs.

### R1–R9 respondent count

```{r}
if (
  sum(
    ess_extended$essround <= 9
  ) !=
    nrow(ess_base_raw)
) {

  stop(
    "R1-R9 respondent count changed during extension."
  )
}
```

### R10–R11 respondent count

```{r}
expected_post9_n <-
  nrow(ess10_raw) +
  nrow(ess11_raw)

observed_post9_n <-
  sum(
    ess_extended$essround >= 10
  )

if (
  expected_post9_n !=
    observed_post9_n
) {

  stop(
    "R10/R11 respondent count changed during extension."
  )
}
```

### Respondent-ID uniqueness

Respondent IDs are expected to be unique within each country-round.

```{r}
duplicate_ids <- ess_extended |>
  count(
    cntry,
    essround,
    idno
  ) |>
  filter(
    n > 1
  )

if (
  nrow(duplicate_ids) > 0
) {

  print(
    duplicate_ids,
    n = Inf
  )

  stop(
    "Duplicate respondent IDs found within country-rounds."
  )
}
```

A compact integrity summary:

```{r}
tibble(
  check = c(
    "R1-R9 respondents",
    "R10-R11 respondents",
    "Duplicate respondent IDs"
  ),
  observed = c(
    sum(ess_extended$essround <= 9),
    observed_post9_n,
    nrow(duplicate_ids)
  ),
  expected = c(
    nrow(ess_base_raw),
    expected_post9_n,
    0
  )
)
```

## 10. Build the extended macro-context file

The project keeps macroeconomic controls separate from the respondent-level dataset.

This preserves a clean distinction between:

- respondent-level ESS data;
- inequality exposure; and
- additional country-round macro controls.

### R1–R9 macro controls

```{r}
macro_base <- read_csv(
  here(
    "03_data",
    "interim",
    "ess_country_round_macro.csv"
  ),
  show_col_types = FALSE
) |>
  select(
    cntry,
    essround,
    log_gdp_pc_ppp_lag1,
    unemployment_lag1
  )
```

### R10–R11 macro controls

```{r}
macro_post9 <- post9_context |>
  select(
    cntry,
    essround,
    log_gdp_pc_ppp_lag1,
    unemployment_lag1
  )
```

### Combine the periods

```{r}
macro_extended <- bind_rows(
  macro_base,
  macro_post9
) |>
  arrange(
    essround,
    cntry
  )
```

Verify uniqueness:

```{r}
macro_duplicates <- macro_extended |>
  count(
    cntry,
    essround
  ) |>
  filter(
    n > 1
  )

if (
  nrow(macro_duplicates) > 0
) {

  print(
    macro_duplicates,
    n = Inf
  )

  stop(
    "Duplicate country-rounds found in extended macro data."
  )
}
```

## 11. Build the combined country-round context file

The combined context object contains one row per country-round with:

- lagged SWIID Gini;
- lagged GDP per capita;
- lagged unemployment; and
- explicit completeness indicators.

```{r}
context_extended <- ess_extended |>
  distinct(
    cntry,
    essround,
    gini_swiid_lag1
  ) |>
  left_join(
    macro_extended,
    by = c(
      "cntry",
      "essround"
    )
  ) |>
  mutate(
    complete_gini =
      !is.na(
        gini_swiid_lag1
      ),

    complete_macro =
      !is.na(
        log_gdp_pc_ppp_lag1
      ) &
      !is.na(
        unemployment_lag1
      ),

    complete_primary_context =
      complete_gini &
      complete_macro
  ) |>
  arrange(
    essround,
    cntry
  )
```

These completeness indicators are useful because later models can identify whether sample changes arise from missing inequality, missing macro controls, or both.

## 12. Respondent-level diagnostics

The extended dataset is summarised by ESS round to inspect:

- respondent and country counts;
- valid social-trust observations;
- missingness in age, gender, education, weights, and inequality;
- mean social trust.

```{r}
build_summary <- ess_extended |>
  group_by(
    essround
  ) |>
  summarise(
    n_respondents =
      n(),

    n_countries =
      n_distinct(
        cntry
      ),

    trust_valid =
      sum(
        !is.na(
          ppltrst
        )
      ),

    trust_missing_share =
      mean(
        is.na(
          ppltrst
        )
      ),

    age_missing_share =
      mean(
        is.na(
          age
        )
      ),

    gender_missing_share =
      mean(
        is.na(
          gender
        )
      ),

    education_missing_share =
      mean(
        is.na(
          education_years
        )
      ),

    weight_missing_share =
      mean(
        is.na(
          analysis_weight
        )
      ),

    gini_missing_share =
      mean(
        is.na(
          gini_swiid_lag1
        )
      ),

    mean_trust =
      mean(
        ppltrst,
        na.rm = TRUE
      ),

    .groups = "drop"
  )

build_summary
```

These diagnostics are not merely descriptive. They are intended to expose discontinuities between the baseline and post-R9 data before the extended dataset enters substantive models.

## 13. Contextual-data diagnostics

Context completeness is summarised separately for:

- R1–R9;
- R10;
- R11.

```{r}
context_summary <- context_extended |>
  mutate(
    period =
      case_when(
        essround <= 9 ~
          "R1-R9",

        essround == 10 ~
          "R10",

        essround == 11 ~
          "R11"
      )
  ) |>
  group_by(
    period
  ) |>
  summarise(
    n_country_rounds =
      n(),

    gini_complete =
      sum(
        complete_gini
      ),

    macro_complete =
      sum(
        complete_macro
      ),

    primary_context_complete =
      sum(
        complete_primary_context
      ),

    primary_context_share =
      mean(
        complete_primary_context
      ),

    .groups = "drop"
  )

context_summary
```

The separation by period is deliberate. It allows later sensitivity analyses to distinguish genuine contextual-data gaps from changes associated with the post-R9 extension.

## 14. Interview-mode diagnostic

Interview mode is especially important in the post-R9 period.

Rather than assuming the later ESS rounds are fully comparable by mode, the harmonised dataset retains both:

- `interview_mode`: standardised mode categories;
- `interview_mode_raw`: original ESS mode label.

The current distribution is summarised here:

```{r}
mode_summary <- ess_extended |>
  filter(
    essround >= 10
  ) |>
  count(
    essround,
    interview_mode,
    name = "n_respondents"
  ) |>
  group_by(
    essround
  ) |>
  mutate(
    share =
      n_respondents /
      sum(n_respondents)
  ) |>
  ungroup()

mode_summary
```

This variable is retained for later robustness and sensitivity analyses rather than automatically entering the primary model specification.

## 15. Save the extended analysis and diagnostic files

Create output directories if necessary:

```{r}
dir.create(
  here(
    "03_data",
    "processed"
  ),
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  here(
    "03_data",
    "interim"
  ),
  recursive = TRUE,
  showWarnings = FALSE
)
```

### Respondent-level extended dataset

```{r}
saveRDS(
  ess_extended,
  here(
    "03_data",
    "processed",
    "ess_analysis_r1_r11.rds"
  )
)
```

### Extended macro and contextual data

```{r}
write_csv(
  macro_extended,
  here(
    "03_data",
    "interim",
    "ess_country_round_macro_r1_r11.csv"
  )
)

write_csv(
  context_extended,
  here(
    "03_data",
    "interim",
    "ess_country_round_context_r1_r11.csv"
  )
)
```

### Diagnostics

```{r}
write_csv(
  build_summary,
  here(
    "03_data",
    "interim",
    "ess_r1_r11_build_summary.csv"
  )
)

write_csv(
  context_summary,
  here(
    "03_data",
    "interim",
    "ess_r1_r11_context_summary.csv"
  )
)

write_csv(
  mode_summary,
  here(
    "03_data",
    "interim",
    "ess_r1_r11_mode_summary.csv"
  )
)
```

## 16. Final build summary

```{r}
final_build_summary <- tibble(
  measure = c(
    "Total respondents",
    "Countries",
    "Country-rounds"
  ),
  value = c(
    nrow(ess_extended),
    n_distinct(
      ess_extended$cntry
    ),
    nrow(context_extended)
  )
)

final_build_summary
```

## Summary of methodological decisions

This extension encodes several important design choices:

1. The R1–R9 dataset remains the baseline. The existing `ess_analysis.rds` file is not overwritten. The R1–R11 dataset is a distinct output.

2. Rounds 10 and 11 are harmonised separately before appending. This keeps later-round transformations explicit and prevents post-R9 changes from being hidden inside the original cleaning pipeline.

3. The respondent-level structure is standardised before `bind_rows()`. Equivalent concepts are given the same names and types across periods.

4. Interview mode is retained explicitly. Later analyses can test whether post-R9 findings are sensitive to interview mode rather than assuming mode equivalence.

5. Trust and demographic variables are range-checked. Unexpected numeric or special codes are converted to missing rather than treated as substantive responses.

6. Analysis weights prioritise `ANWEIGHT`. When it is unavailable, the script reconstructs the weight from `PSPWGHT × PWEIGHT`.

7. **Lagged SWIID inequality is attached to R10–R11 through a country-round context file. Respondents remain the backbone of the join, so missing context does not silently drop cases.

8. Macroeconomic controls remain in a separate country-round file. This preserves a clean distinction between respondent data and contextual covariates.

9. The post-R9 period remains analytically identifiable. `post_r9` and `period` allow R1–R9, R10, and R11 to be compared directly.

10. Integrity checks are treated as part of the build process. The script stops if respondent counts change unexpectedly, respondent IDs duplicate within country-rounds, or contextual files contain duplicate country-rounds.

11. Context completeness is recorded explicitly. Later models can identify whether sample restrictions arise from missing Gini, missing macro controls, or both.

12. Diagnostic summaries are saved alongside the data. This creates an audit trail for respondent-level missingness, context completeness, and interview mode.

## Next steps

The extended R1–R11 dataset provides the respondent-level backbone for analyses that go beyond the original R1–R9 period.

Subsequent work can use it to:

- estimate the original inequality–trust models on R1–R9 and then extend them to R10–R11;
- model period interactions explicitly;
- examine whether R10 behaves differently from R11;
- conduct interview-mode sensitivity analyses; and
- merge additional respondent-level political variables and party-linked discourse measures once those constructions are complete.
