---
title: "Propagating SWIID inequality uncertainty"
subtitle: "Input audit, imputation-specific exposure construction and Rubin pooling"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-copy: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Purpose and status

This notebook documents the treatment of uncertainty in the Standardized World Income Inequality Database (SWIID) version 9.92.

Two executable scripts underpin the workflow:

```text
04_code/02_descriptives/10_swiid_uncertainty_audit.R
04_code/03_models/10_swiid_uncertainty.R
```

The first audits the structure and scale of the released SWIID imputations.

The second propagates those imputations through the Rounds 1–9 random-effects within-between (REWB) models.

The modelling workflow was first validated using three SWIID imputations and has now been completed for all 100 imputations in the Rounds 1–9 sample.

Across the full run, the analytical samples were identical across imputations, all 200 fitted multilevel models converged without singularity warnings, and the Rubin-pooled results led to the same substantive conclusions as the models based on the SWIID summary series.

The next step is to extend the same uncertainty-propagation workflow to the Rounds 1–11 sample and the Rounds 1–11 in-person-only sensitivity analysis.

This notebook does **not** refit the multilevel models when rendered. It reads the saved diagnostic and model outputs produced by the executable scripts.

## 1. Why propagate SWIID uncertainty?

The primary inequality measure is the SWIID disposable-income Gini coefficient.

SWIID estimates are model-based and therefore contain measurement uncertainty.

The main summary-series models treat the reported country-year Gini values as observed contextual covariates.

The uncertainty workflow asks whether the substantive conclusions change when the uncertainty represented by the released SWIID imputations is propagated through:

1. country-round exposure construction;
2. within-between decomposition;
3. model estimation;
4. coefficient pooling.

The purpose is not to replace the summary-series analysis automatically.

It is to assess how much the uncertainty in the inequality exposure contributes to uncertainty in the estimated inequality-trust relationship.

## 2. SWIID input structure

The R-format SWIID 9.92 file is:

```text
03_data/raw/inequality/swiid/swiid9_92.rda
```

It contains two relevant objects:

- `swiid`: a list of 100 imputed country-year inequality datasets;
- `swiid_summary`: the published country-year summary values and associated uncertainty.

The uncertainty audit verifies:

- exactly 100 imputations are present;
- each imputation has the expected structure;
- `country`, `year`, and `gini_disp` are present;
- country-year keys are unique within each imputation;
- the country-year structure of the imputations matches `swiid_summary`.

## 3. Gini scale

The two SWIID objects use different numerical representations.

In `swiid_summary`, `gini_disp` is already on the conventional 0–100 Gini scale.

In the individual imputation objects, the raw values are stored in hundredths.

Thus:

$$
Gini^{scaled}_{cy,m}
=
\frac{Gini^{raw}_{cy,m}}{100},
$$

where:

- $c$ indexes country;
- $y$ indexes year;
- $m$ indexes SWIID imputation.

This scale conversion is required before constructing ESS country-round exposures.

The audit stops if the raw values do not have the expected scale.

## 4. Descriptive comparison with `swiid_summary`

The uncertainty audit calculates, for every country-year:

- the empirical mean across the 100 released imputations;
- the empirical standard deviation across the 100 released imputations.

These quantities are compared descriptively with the published `swiid_summary` point estimate and uncertainty measure.

The comparison is a quality-assurance check.

It does **not** assume that:

- the 100 released imputations are independent Monte Carlo draws whose empirical mean must reproduce `swiid_summary`;
- `gini_disp_se / sqrt(100)` is an appropriate tolerance for the difference between the imputation mean and the published summary estimate.

No formal Monte Carlo consistency test is therefore applied.

The released imputations and the summary series need not reproduce one another exactly.

## 5. Saved audit diagnostics

The audit writes compact outputs to:

```text
05_output/tables/swiid_uncertainty_audit/
```

```{r}
#| label: setup

library(dplyr)
library(here)
library(readr)

read_optional <- function(path) {
  if (file.exists(path)) {
    read_csv(
      path,
      show_col_types = FALSE
    )
  } else {
    tibble()
  }
}

audit_dir <- here(
  "05_output",
  "tables",
  "swiid_uncertainty_audit"
)

model_dir <- here(
  "05_output",
  "tables",
  "swiid_uncertainty"
)
```

### Imputation coverage

```{r}
#| label: swiid-imputation-coverage

imputation_coverage <- read_optional(
  file.path(
    audit_dir,
    "swiid_imputation_coverage.csv"
  )
)

imputation_coverage
```

The completed audit found complete 100-imputation coverage for each of the 6,628 SWIID country-years represented in both the imputation data and `swiid_summary`.

### Country-year structure

```{r}
#| label: swiid-structure

structure_check <- read_optional(
  file.path(
    audit_dir,
    "swiid_summary_draw_structure.csv"
  )
)

structure_check
```

The expected result is:

- no country-years present only in `swiid_summary`;
- no country-years present only in the imputation data;
- 6,628 country-years present in both.

### Descriptive agreement

```{r}
#| label: swiid-summary-comparison

agreement <- read_optional(
  file.path(
    audit_dir,
    "swiid_summary_draw_agreement.csv"
  )
)

agreement
```

The purpose of this table is descriptive.

Differences between the published summary values and empirical moments of the released imputations are reported rather than judged against a formal Monte Carlo tolerance.

## 6. ESS timing and exposure construction

The uncertainty models use the same country-round timing logic as the main analysis.

For each country-round, the observed ESS interview-year distribution is used to construct a one-year-lagged Gini exposure.

For SWIID imputation $m$:

$$
Gini^{lag1}_{ct,m}
=
\sum_y
s_{cty}
Gini_{c,y-1,m},
$$

where $s_{cty}$ is the share of dated ESS respondents in country $c$, round $t$, interviewed in year $y$.

A crucial feature is that all annual values entering one country-round exposure come from the **same SWIID imputation**.

Annual values from different imputations are never mixed within one contextual exposure.

## 7. Complete-component rule

As in the main contextual-data workflow, an imputation-specific country-round Gini is constructed only if all annual components required by the observed ESS timing are available.

The model script does not renormalise the interview-year weights over available SWIID years.

If a country-round requires two lagged annual values and one is missing in a given imputation, the country-round Gini for that imputation is missing.

This keeps the uncertainty workflow consistent with the main exposure-construction rule.

## 8. Recalculate the within-between decomposition within each imputation

The Gini decomposition is recalculated separately for every SWIID imputation.

This is necessary because uncertainty in the annual inequality values affects both:

- the country's estimated average inequality;
- the country-round deviation from that average.

For country $c$, round $t$, and imputation $m$:

$$
Gini^{B}_{c,m}
=
\frac{1}{T_c}
\sum_t
Gini_{ct,m},
$$

and:

$$
Gini^{W}_{ct,m}
=
Gini_{ct,m}
-
Gini^{B}_{c,m}.
$$

The decomposition is calculated from unique country-round observations.

Respondent sample size therefore does not give larger ESS country-rounds greater weight when calculating country-level contextual means.

The between-country component is grand-mean centred before model estimation.

## 9. M3 and M4 samples

The uncertainty workflow fits two established Rounds 1–9 specifications.

### M3

M3 includes:

- within-country Gini;
- between-country Gini;
- ESS round fixed effects;
- age and age squared;
- gender;
- education;
- country random intercept;
- country-round random intercept.

In schematic form:

$$
Trust_{ict}
=
\beta_0
+
\beta_W Gini^W_{ct}
+
\beta_B Gini^B_c
+
\mathbf{X}_{ict}\boldsymbol{\gamma}
+
\delta_t
+
u_c
+
v_{ct}
+
\epsilon_{ict}.
$$

### M4

M4 adds within-between decompositions of:

- log GDP per capita, PPP;
- unemployment.

The contextual decomposition for M4 is recalculated on the complete macro-data sample for each imputation.

This is important because the relevant country means should correspond to the observations actually contributing to the fitted model.

Both models are estimated using maximum likelihood with the same random-effects structure and optimiser settings as the corresponding main REWB analysis.

No survey weights are currently applied, matching the primary models.

## 10. Sample stability across imputations

SWIID uncertainty should alter the Gini values, not arbitrarily alter which ESS observations enter the model.

The script therefore checks that, within each model, all fitted imputations have identical:

- respondent counts;
- country counts;
- country-round counts.

```{r}
#| label: uncertainty-sample-stability

sample_stability <- read_optional(
  file.path(
    model_dir,
    "swiid_sample_stability.csv"
  )
)

sample_stability
```

Across all 100 SWIID imputations, the samples were completely stable:

| Model | Respondents | Countries | Country-rounds |
|---|---:|---:|---:|
| M3 | 420,905 | 38 | 227 |
| M4 | 419,618 | 37 | 226 |

The script stops if those counts vary across imputations. No sample variation occurred in the completed 100-imputation run.

## 11. Model diagnostics

Each fitted model is checked for:

- singularity;
- optimiser convergence messages.

```{r}
#| label: uncertainty-model-diagnostics

model_diagnostics <- read_optional(
  file.path(
    model_dir,
    "swiid_model_diagnostics.csv"
  )
)

model_diagnostics
```

The initial three-imputation validation produced six fitted models and no singular fits or convergence messages.

The completed 100-imputation analysis produced 200 fitted models: 100 for M3 and 100 for M4.

None of the models was singular and none produced an optimiser convergence message.

This indicates that the uncertainty-propagation workflow is computationally stable across the full set of released SWIID imputations.

## 12. Coefficient variation across imputations

Before pooling, the workflow summarises how the Gini coefficients vary across imputations.

```{r}
#| label: gini-imputation-variation

gini_variation <- read_optional(
  file.path(
    model_dir,
    "swiid_gini_imputation_summary.csv"
  )
)

gini_variation
```

This diagnostic helps distinguish two situations:

1. SWIID uncertainty produces very little variation in the estimated coefficient;
2. SWIID uncertainty is large enough to contribute materially to inferential uncertainty.

The comparison should be made relative to the model-based standard errors, not from the raw coefficient range alone.

## 13. Rubin pooling

Fixed-effect estimates are pooled using Rubin's rules.

For coefficient $Q$, let:

$$
\bar{Q}
=
\frac{1}{m}
\sum_{j=1}^{m}
Q_j
$$

be the average coefficient across $m$ imputations.

The average within-imputation sampling variance is:

$$
\bar{U}
=
\frac{1}{m}
\sum_{j=1}^{m}
U_j.
$$

The between-imputation variance is:

$$
B
=
\frac{1}{m-1}
\sum_{j=1}^{m}
(Q_j-\bar{Q})^2.
$$

The total variance is:

$$
T
=
\bar{U}
+
\left(
1+\frac{1}{m}
\right)B.
$$

The pooled standard error is:

$$
SE(\bar{Q})
=
\sqrt{T}.
$$

This combines:

- ordinary model-based sampling uncertainty;
- additional uncertainty associated with the SWIID inequality imputations.

The script also calculates an approximate Rubin degrees of freedom and the fraction of missing information.

## 14. Fraction of missing information

For the implemented pooling calculation, the fraction of missing information is based on the share of total coefficient variance attributable to between-imputation uncertainty:

$$
FMI
=
\frac{
\left(1+\frac{1}{m}\right)B
}{
T
}.
$$

A small value indicates that SWIID imputation uncertainty contributes relatively little to total uncertainty for that coefficient.

A larger value indicates that the inequality measurement uncertainty has a more material effect on inference.

The substantive interpretation should focus especially on the within-country Gini coefficient because it is the main inequality parameter of interest.

## 15. Pooled Gini results

```{r}
#| label: pooled-gini-results

pooled_gini <- read_optional(
  file.path(
    model_dir,
    "swiid_pooled_gini_results.csv"
  )
)

pooled_gini
```


```{r}
#| label: uncertainty-run-status

run_status <- if (
  nrow(sample_stability) > 0
) {
  sample_stability |>
    summarise(
      imputations_run =
        max(n_imputations),
      status =
        if_else(
          imputations_run == 100,
          "Full 100-imputation run",
          "Validation run only"
        )
    )
} else {
  tibble(
    imputations_run = NA_integer_,
    status = "No saved model outputs found"
  )
}

run_status
```

## 16. Comparison with the summary-series analysis

Where the existing point-estimate results are available, the workflow compares the SWIID-pooled Gini coefficients with the models based on `swiid_summary`.

```{r}
#| label: point-vs-pooled

point_vs_pooled <- read_optional(
  file.path(
    model_dir,
    "swiid_point_vs_pooled_gini.csv"
  )
)

point_vs_pooled
```

The comparison reports:

- the change in the coefficient estimate;
- the change in its standard error;
- the pooled confidence interval;
- the fraction of missing information.

Two distinct questions should be kept separate:

1. Does propagating SWIID uncertainty materially change the point estimate?
2. Does it materially increase uncertainty around that estimate?

A stable point estimate can still be accompanied by a wider uncertainty interval.

## 17. Why the summary-series audit is descriptive

The audit deliberately avoids treating the empirical mean of the 100 released SWIID imputations as a Monte Carlo estimate that must equal `swiid_summary`.

In particular, the workflow does not standardise:

$$
\bar{Gini}_{draws}
-
Gini_{summary}
$$

using:

$$
\frac{SE_{summary}}{\sqrt{100}}.
$$

That diagnostic was considered and rejected because it imposes an unverified interpretation of the relationship between the released imputations and the published summary object.

The retained audit therefore checks:

- structural consistency;
- scale;
- country-year coverage;
- descriptive agreement.

The inferential use of the imputations occurs in the model pooling stage, where uncertainty is propagated through the actual estimand of interest.

## 18. Completed Rounds 1–9 uncertainty analysis

The full 100-imputation Rounds 1–9 analysis has now been completed.

The pooled estimates for the Gini coefficients were:

| Model | Effect | Pooled estimate | SE | 95% CI | p | FMI |
|---|---|---:|---:|---:|---:|---:|
| M3 | Between-country Gini | -0.134 | 0.0330 | [-0.198, -0.069] | < .001 | 0.038 |
| M3 | Within-country Gini | 0.0018 | 0.0139 | [-0.025, 0.029] | .896 | 0.412 |
| M4 | Between-country Gini | -0.0465 | 0.0283 | [-0.102, 0.009] | .101 | 0.036 |
| M4 | Within-country Gini | 0.0070 | 0.0134 | [-0.019, 0.033] | .605 | 0.410 |

Propagating SWIID uncertainty therefore does not materially change the substantive conclusions of the Rounds 1–9 models.

For the within-country Gini coefficient, the pooled estimates remain close to zero and statistically imprecise in both M3 and M4.

The fraction of missing information is approximately 0.41 for the within-country Gini coefficient in both models, indicating that variation across SWIID imputations contributes meaningfully to uncertainty in this parameter.

However, the resulting pooled standard errors are very similar to those from the corresponding summary-series models:

| Model | Summary-series SE | Pooled SE |
|---|---:|---:|
| M3 within-country Gini | 0.0136 | 0.0139 |
| M4 within-country Gini | 0.0133 | 0.0134 |

Thus, although SWIID measurement uncertainty is non-negligible, propagating it through the models does not materially alter either the estimated within-country inequality effect or the overall inferential conclusion.

The between-country estimates are even less sensitive to SWIID uncertainty, with fractions of missing information below 0.04 in both M3 and M4.

The M3 between-country association remains strongly negative after pooling, whereas the corresponding association is substantially attenuated after adjustment for GDP and unemployment in M4.

## 19. Planned extensions

Having completed the full Rounds 1–9 uncertainty analysis, the same logic should now be extended to:

1. the Rounds 1–11 analysis;
2. the Rounds 1–11 in-person-only sensitivity analysis.

The central design principles should remain unchanged:

- use the same imputation consistently across all annual components of one exposure;
- reconstruct country-round exposures within each imputation;
- recalculate within-between components within each imputation;
- preserve the complete-component rule;
- fit the same substantive specification within each imputation;
- pool fixed-effect estimates only after model fitting.

## 20. Reproducibility and interpretation

The executable scripts remain authoritative.

The audit script verifies the SWIID inputs before modelling.

The model script produces the imputation-specific models and Rubin-pooled outputs.

This notebook is intended to make transparent the non-obvious methodological decisions connecting those two stages.

The uncertainty analysis should be interpreted as a sensitivity analysis for uncertainty in the SWIID inequality exposure.

It does not address:

- uncertainty in V-Dem political-discourse measures;
- uncertainty in World Bank macroeconomic controls;
- alternative survey-weighting estimands;
- broader model-specification uncertainty.

Those issues require separate sensitivity analyses.