---
title: "Propagating SWIID inequality uncertainty"
subtitle: "Input audit, imputation-specific exposure construction and Rubin pooling"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-copy: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Purpose and status

This notebook documents the treatment of uncertainty in the Standardized World Income Inequality Database (SWIID) version 9.92.

Four executable scripts underpin the workflow:

```text
04_code/02_descriptives/10_swiid_uncertainty_audit.R
04_code/03_models/10_swiid_uncertainty.R
04_code/03_models/11_swiid_uncertainty_r1_r11.R
04_code/03_models/12_swiid_uncertainty_r1_r11_inperson.R
```

The first audits the structure and scale of the released SWIID imputations.

The second propagates all 100 SWIID imputations through the Rounds 1–9 random-effects within-between (REWB) models.

The third applies the same uncertainty-propagation workflow to the full Rounds 1–11 sample.

The fourth validates the workflow after restricting the respondent sample to in-person interviews. This final check uses three SWIID imputations rather than repeating the full 100-imputation analysis.

The full 100-imputation analyses have now been completed for both Rounds 1–9 and Rounds 1–11. In both samples:

- analytical samples were identical across all 100 imputations;
- all fitted multilevel models converged without singularity warnings;
- Rubin-pooled estimates closely reproduced the conclusions from models based on the SWIID summary series.

The Rounds 1–11 in-person sensitivity was additionally checked across three SWIID imputations. The analytical sample was stable, all fitted models converged cleanly, and the imputation-specific and pooled coefficients remained close to the corresponding summary-series in-person estimates.

Because the full Rounds 1–11 uncertainty analysis had already established robustness to SWIID measurement uncertainty, the in-person analysis was treated as a validation of the mode-restricted specification rather than as a second full 100-imputation uncertainty analysis.

This notebook does not refit the multilevel models when rendered. It reads the saved diagnostic and model outputs produced by the executable scripts.

## 1. Why propagate SWIID uncertainty?

The primary inequality measure is the SWIID disposable-income Gini coefficient.

SWIID estimates are model-based and therefore contain measurement uncertainty.

The main summary-series models treat the reported country-year Gini values as observed contextual covariates.

The uncertainty workflow asks whether the substantive conclusions change when the uncertainty represented by the released SWIID imputations is propagated through:

1. country-round exposure construction;
2. within-between decomposition;
3. model estimation;
4. coefficient pooling.

The purpose is not to replace the summary-series analysis automatically.

It is to assess how much the uncertainty in the inequality exposure contributes to uncertainty in the estimated inequality-trust relationship.

## 2. SWIID input structure

The R-format SWIID 9.92 file is:

```text
03_data/raw/inequality/swiid/swiid9_92.rda
```

It contains two relevant objects:

- `swiid`: a list of 100 imputed country-year inequality datasets;
- `swiid_summary`: the published country-year summary values and associated uncertainty.

The uncertainty audit verifies:

- exactly 100 imputations are present;
- each imputation has the expected structure;
- `country`, `year`, and `gini_disp` are present;
- country-year keys are unique within each imputation;
- the country-year structure of the imputations matches `swiid_summary`.

## 3. Gini scale

The two SWIID objects use different numerical representations.

In `swiid_summary`, `gini_disp` is already on the conventional 0–100 Gini scale.

In the individual imputation objects, the raw values are stored in hundredths.

Thus:

$$
Gini^{scaled}_{cy,m}
=
\frac{Gini^{raw}_{cy,m}}{100},
$$

where:

- $c$ indexes country;
- $y$ indexes year;
- $m$ indexes SWIID imputation.

This scale conversion is required before constructing ESS country-round exposures.

The audit stops if the raw values do not have the expected scale.

## 4. Descriptive comparison with `swiid_summary`

The uncertainty audit calculates, for every country-year:

- the empirical mean across the 100 released imputations;
- the empirical standard deviation across the 100 released imputations.

These quantities are compared descriptively with the published `swiid_summary` point estimate and uncertainty measure.

The comparison is a quality-assurance check.

It does **not** assume that:

- the 100 released imputations are independent Monte Carlo draws whose empirical mean must reproduce `swiid_summary`;
- `gini_disp_se / sqrt(100)` is an appropriate tolerance for the difference between the imputation mean and the published summary estimate.

No formal Monte Carlo consistency test is therefore applied.

The released imputations and the summary series need not reproduce one another exactly.

## 5. Saved audit diagnostics

The audit writes compact outputs to:

```text
05_output/tables/swiid_uncertainty_audit/
```

```{r}
#| label: setup

library(dplyr)
library(here)
library(readr)

read_optional <- function(path) {
  if (file.exists(path)) {
    read_csv(
      path,
      show_col_types = FALSE
    )
  } else {
    tibble()
  }
}

audit_dir <- here(
  "05_output",
  "tables",
  "swiid_uncertainty_audit"
)

model_dir <- here(
  "05_output",
  "tables",
  "swiid_uncertainty"
)

model_dir_r1_r11 <- here(
  "05_output",
  "tables",
  "swiid_uncertainty_r1_r11"
)

model_dir_inperson <- here(
  "05_output",
  "tables",
  "swiid_uncertainty_r1_r11_inperson"
)
```

### Imputation coverage

```{r}
#| label: swiid-imputation-coverage

imputation_coverage <- read_optional(
  file.path(
    audit_dir,
    "swiid_imputation_coverage.csv"
  )
)

imputation_coverage
```

The completed audit found complete 100-imputation coverage for each of the 6,628 SWIID country-years represented in both the imputation data and `swiid_summary`.

### Country-year structure

```{r}
#| label: swiid-structure

structure_check <- read_optional(
  file.path(
    audit_dir,
    "swiid_summary_draw_structure.csv"
  )
)

structure_check
```

The expected result is:

- no country-years present only in `swiid_summary`;
- no country-years present only in the imputation data;
- 6,628 country-years present in both.

### Descriptive agreement

```{r}
#| label: swiid-summary-comparison

agreement <- read_optional(
  file.path(
    audit_dir,
    "swiid_summary_draw_agreement.csv"
  )
)

agreement
```

The purpose of this table is descriptive.

Differences between the published summary values and empirical moments of the released imputations are reported rather than judged against a formal Monte Carlo tolerance.

## 6. ESS timing and exposure construction

The uncertainty models use the same country-round timing logic as the main analysis.

For each country-round, the observed ESS interview-year distribution is used to construct a one-year-lagged Gini exposure.

For SWIID imputation $m$:

$$
Gini^{lag1}_{ct,m}
=
\sum_y
s_{cty}
Gini_{c,y-1,m},
$$

where $s_{cty}$ is the share of dated ESS respondents in country $c$, round $t$, interviewed in year $y$.

A crucial feature is that all annual values entering one country-round exposure come from the **same SWIID imputation**.

Annual values from different imputations are never mixed within one contextual exposure.

## 7. Complete-component rule

As in the main contextual-data workflow, an imputation-specific country-round Gini is constructed only if all annual components required by the observed ESS timing are available.

The model script does not renormalise the interview-year weights over available SWIID years.

If a country-round requires two lagged annual values and one is missing in a given imputation, the country-round Gini for that imputation is missing.

This keeps the uncertainty workflow consistent with the main exposure-construction rule.

## 8. Recalculate the within-between decomposition within each imputation

The Gini decomposition is recalculated separately for every SWIID imputation.

This is necessary because uncertainty in the annual inequality values affects both:

- the country's estimated average inequality;
- the country-round deviation from that average.

For country $c$, round $t$, and imputation $m$:

$$
Gini^{B}_{c,m}
=
\frac{1}{T_c}
\sum_t
Gini_{ct,m},
$$

and:

$$
Gini^{W}_{ct,m}
=
Gini_{ct,m}
-
Gini^{B}_{c,m}.
$$

The decomposition is calculated from unique country-round observations.

Respondent sample size therefore does not give larger ESS country-rounds greater weight when calculating country-level contextual means.

The between-country component is grand-mean centred before model estimation.

## 9. M3 and M4 samples

The uncertainty workflow evaluates the established M3 and M4 specifications in three analytical samples:

1. Rounds 1–9;
2. Rounds 1–11;
3. Rounds 1–11 restricted to in-person interviews.

### M3

M3 includes:

- within-country Gini;
- between-country Gini;
- ESS round fixed effects;
- age and age squared;
- gender;
- education;
- country random intercept;
- country-round random intercept.

In schematic form:

$$
Trust_{ict}
=
\beta_0
+
\beta_W Gini^W_{ct}
+
\beta_B Gini^B_c
+
\mathbf{X}_{ict}\boldsymbol{\gamma}
+
\delta_t
+
u_c
+
v_{ct}
+
\epsilon_{ict}.
$$

### M4

M4 adds within-between decompositions of:

- log GDP per capita, PPP;
- unemployment.

The contextual decomposition for M4 is recalculated on the complete macro-data sample within each SWIID imputation.

Both models are estimated using maximum likelihood with the same random-effects structure and optimiser settings as the corresponding main REWB analysis.

No survey weights are currently applied, matching the primary models.

### In-person sensitivity

For the Rounds 1–11 in-person sensitivity, the contextual inequality exposure is **not** reconstructed using only the interview dates of in-person respondents.

Instead, the same country-round exposure definition used in the main Rounds 1–11 analysis is retained, based on the full country-round distribution of interview years.

Only the respondent-level estimation sample is restricted to:

```text
interview_mode == "In-person"
```

This keeps the contextual exposure fixed and isolates the interview-mode restriction rather than changing both survey mode and exposure construction simultaneously.

## 10. Sample stability across imputations

SWIID uncertainty should alter the Gini values, not arbitrarily alter which ESS observations enter the model.

The scripts therefore check that all fitted imputations within each analysis have identical:

- respondent counts;
- country counts;
- country-round counts.

```{r}
#| label: uncertainty-sample-stability

sample_stability <- bind_rows(
  read_optional(
    file.path(
      model_dir,
      "swiid_sample_stability.csv"
    )
  ) |>
    mutate(
      sample = "R1-R9",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_r1_r11,
      "swiid_sample_stability.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_inperson,
      "swiid_sample_stability.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11 in-person",
      .before = 1
    )
)

sample_stability
```

The resulting samples were completely stable within each analysis:

| Sample | Model | Respondents | Countries | Country-rounds | Imputations |
|---|---|---:|---:|---:|---:|
| R1–R9 | M3 | 420,905 | 38 | 227 | 100 |
| R1–R9 | M4 | 419,618 | 37 | 226 | 100 |
| R1–R11 | M3 | 499,748 | 39 | 274 | 100 |
| R1–R11 | M4 | 498,461 | 38 | 273 | 100 |
| R1–R11 in-person | M3 | 496,277 | 39 | 274 | 3 |
| R1–R11 in-person | M4 | 494,990 | 38 | 273 | 3 |

The in-person restriction therefore reduces the respondent count but does not reduce the number of usable countries or country-rounds relative to the corresponding all-mode Rounds 1–11 models.

## 11. Model diagnostics

Each fitted model is checked for:

- singularity;
- optimiser convergence messages.

```{r}
#| label: uncertainty-model-diagnostics

model_diagnostics <- bind_rows(
  read_optional(
    file.path(
      model_dir,
      "swiid_model_diagnostics.csv"
    )
  ) |>
    mutate(
      sample = "R1-R9",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_r1_r11,
      "swiid_model_diagnostics.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_inperson,
      "swiid_model_diagnostics.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11 in-person",
      .before = 1
    )
)

diagnostic_summary <- model_diagnostics |>
  group_by(
    sample,
    model
  ) |>
  summarise(
    n_fits = n(),

    singular_fits =
      sum(
        singular,
        na.rm = TRUE
      ),

    convergence_messages =
      sum(
        !is.na(
          convergence_message
        )
      ),

    .groups = "drop"
  )

diagnostic_summary
```

The completed workflow produced:

- 200 fitted models for Rounds 1–9;
- 200 fitted models for Rounds 1–11;
- six fitted models for the three-imputation Rounds 1–11 in-person validation.

None was singular and none produced an optimiser convergence message.

## 12. Coefficient variation across imputations

Before pooling, the workflow summarises how the Gini coefficients vary across imputations.

```{r}
#| label: gini-imputation-variation

gini_variation <- bind_rows(
  read_optional(
    file.path(
      model_dir,
      "swiid_gini_imputation_summary.csv"
    )
  ) |>
    mutate(
      sample = "R1-R9",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_r1_r11,
      "swiid_gini_imputation_summary.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_inperson,
      "swiid_gini_imputation_summary.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11 in-person",
      .before = 1
    )
)

gini_variation
```

The comparison should be interpreted relative to the model-based standard errors rather than from the raw coefficient range alone.

For the in-person validation, the three-imputation variation is used only as a diagnostic. Three imputations are not sufficient for treating the resulting Rubin variance estimates, confidence intervals or fraction of missing information as final inferential results.

## 13. Rubin pooling

Fixed-effect estimates are pooled using Rubin's rules.

For coefficient $Q$, let:

$$
\bar{Q}
=
\frac{1}{m}
\sum_{j=1}^{m}
Q_j
$$

be the average coefficient across $m$ imputations.

The average within-imputation sampling variance is:

$$
\bar{U}
=
\frac{1}{m}
\sum_{j=1}^{m}
U_j.
$$

The between-imputation variance is:

$$
B
=
\frac{1}{m-1}
\sum_{j=1}^{m}
(Q_j-\bar{Q})^2.
$$

The total variance is:

$$
T
=
\bar{U}
+
\left(
1+\frac{1}{m}
\right)B.
$$

The pooled standard error is:

$$
SE(\bar{Q})
=
\sqrt{T}.
$$

This combines:

- ordinary model-based sampling uncertainty;
- additional uncertainty associated with the SWIID inequality imputations.

The script also calculates an approximate Rubin degrees of freedom and the fraction of missing information.

## 14. Fraction of missing information

For the implemented pooling calculation, the fraction of missing information is based on the share of total coefficient variance attributable to between-imputation uncertainty:

$$
FMI
=
\frac{
\left(1+\frac{1}{m}\right)B
}{
T
}.
$$

A small value indicates that SWIID imputation uncertainty contributes relatively little to total uncertainty for that coefficient.

A larger value indicates that the inequality measurement uncertainty has a more material effect on inference.

The substantive interpretation should focus especially on the within-country Gini coefficient because it is the main inequality parameter of interest.

## 15. Pooled Gini results

```{r}
#| label: pooled-gini-results

pooled_gini <- bind_rows(
  read_optional(
    file.path(
      model_dir,
      "swiid_pooled_gini_results.csv"
    )
  ) |>
    mutate(
      sample = "R1-R9",
      analysis_status = "Full 100-imputation analysis",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_r1_r11,
      "swiid_pooled_gini_results.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11",
      analysis_status = "Full 100-imputation analysis",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_inperson,
      "swiid_pooled_gini_results.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11 in-person",
      analysis_status = "Three-imputation validation only",
      .before = 1
    )
)

pooled_gini
```

The Rounds 1–9 and Rounds 1–11 pooled estimates are based on all 100 released SWIID imputations.

The in-person results are retained for validation only and should not be presented as final Rubin-pooled inference.

## 16. Comparison with the summary-series analysis

The workflow compares the SWIID-based estimates with the corresponding models using the SWIID summary series.

```{r}
#| label: point-vs-pooled

point_vs_pooled <- bind_rows(
  read_optional(
    file.path(
      model_dir,
      "swiid_point_vs_pooled_gini.csv"
    )
  ) |>
    mutate(
      sample = "R1-R9",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_r1_r11,
      "swiid_point_vs_pooled_gini.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11",
      .before = 1
    ),

  read_optional(
    file.path(
      model_dir_inperson,
      "swiid_point_vs_pooled_gini.csv"
    )
  ) |>
    mutate(
      sample = "R1-R11 in-person",
      .before = 1
    )
)

point_vs_pooled
```

Two distinct questions remain important:

1. Does propagating SWIID uncertainty materially change the point estimate?
2. Does it materially increase uncertainty around that estimate?

For the two full 100-imputation analyses, both questions can be evaluated directly.

For the in-person sensitivity, the comparison is used only to establish that the three sampled SWIID draws do not indicate an anomalous dependence of the mode-restricted results on inequality measurement uncertainty.

## 17. Why the summary-series audit is descriptive

The audit deliberately avoids treating the empirical mean of the 100 released SWIID imputations as a Monte Carlo estimate that must equal `swiid_summary`.

In particular, the workflow does not standardise:

$$
\bar{Gini}_{draws}
-
Gini_{summary}
$$

using:

$$
\frac{SE_{summary}}{\sqrt{100}}.
$$

That diagnostic was considered and rejected because it imposes an unverified interpretation of the relationship between the released imputations and the published summary object.

The retained audit therefore checks:

- structural consistency;
- scale;
- country-year coverage;
- descriptive agreement.

The inferential use of the imputations occurs in the model pooling stage, where uncertainty is propagated through the actual estimand of interest.

## 18. Completed uncertainty analyses

### 18.1 Rounds 1–9

The full 100-imputation Rounds 1–9 analysis produced:

| Model | Effect | Pooled estimate | SE | 95% CI | p | FMI |
|---|---|---:|---:|---:|---:|---:|
| M3 | Between-country Gini | -0.134 | 0.0330 | [-0.198, -0.069] | < .001 | 0.038 |
| M3 | Within-country Gini | 0.0018 | 0.0139 | [-0.025, 0.029] | .896 | 0.412 |
| M4 | Between-country Gini | -0.0465 | 0.0283 | [-0.102, 0.009] | .101 | 0.036 |
| M4 | Within-country Gini | 0.0070 | 0.0134 | [-0.019, 0.033] | .605 | 0.410 |

Propagating SWIID uncertainty does not materially alter the Rounds 1–9 conclusions.

For the within-country Gini coefficient, the fraction of missing information is approximately 0.41 in both M3 and M4, indicating meaningful variation across SWIID imputations.

However, the pooled standard errors remain almost identical to the corresponding summary-series standard errors.

| Model | Summary-series SE | Pooled SE |
|---|---:|---:|
| M3 within-country Gini | 0.0136 | 0.0139 |
| M4 within-country Gini | 0.0133 | 0.0134 |

Thus, SWIID measurement uncertainty is non-negligible but does not materially alter inference.

### 18.2 Rounds 1–11

The full 100-imputation Rounds 1–11 analysis produced:

| Model | Effect | Pooled estimate | SE | 95% CI | p | FMI |
|---|---|---:|---:|---:|---:|---:|
| M3 | Between-country Gini | -0.135 | 0.0333 | [-0.200, -0.070] | < .001 | 0.043 |
| M3 | Within-country Gini | 0.0010 | 0.0113 | [-0.021, 0.023] | .929 | 0.370 |
| M4 | Between-country Gini | -0.0447 | 0.0280 | [-0.100, 0.010] | .110 | 0.038 |
| M4 | Within-country Gini | 0.0049 | 0.0112 | [-0.017, 0.027] | .663 | 0.380 |

The corresponding summary-series estimates were:

| Model | Effect | Summary-series estimate | Summary-series SE |
|---|---|---:|---:|
| M3 | Between-country Gini | -0.137 | 0.0327 |
| M3 | Within-country Gini | 0.0008 | 0.0114 |
| M4 | Between-country Gini | -0.0430 | 0.0276 |
| M4 | Within-country Gini | 0.0073 | 0.0112 |

The pooled and summary-series results are therefore extremely similar.

For the within-country Gini coefficients, the fractions of missing information are approximately 0.37–0.38, again showing that SWIID uncertainty contributes meaningfully to the variance decomposition.

Nevertheless, the ratios of pooled to summary-series standard errors are approximately one:

| Model | SE ratio |
|---|---:|
| M3 within-country Gini | 0.994 |
| M4 within-country Gini | 0.997 |

Explicit propagation of SWIID uncertainty therefore has essentially no effect on the overall precision or substantive interpretation of the Rounds 1–11 within-country inequality estimates.

The results also closely reproduce the Rounds 1–9 findings. Extending the analysis through Rounds 10–11 does not materially alter the average within-country or between-country Gini coefficients.

### 18.3 Rounds 1–11 in-person validation

The interview-mode sensitivity restricts the respondent sample to in-person interviews while retaining the same country-round contextual exposure definition used in the main Rounds 1–11 analysis.

The summary-series in-person estimates were compared with estimates from three SWIID imputations:

| Model | Effect | Summary-series estimate | Three-draw pooled estimate | Difference |
|---|---|---:|---:|---:|
| M3 | Between-country Gini | -0.137 | -0.134 | 0.0025 |
| M3 | Within-country Gini | 0.0016 | 0.0006 | -0.0011 |
| M4 | Between-country Gini | -0.0431 | -0.0425 | 0.0006 |
| M4 | Within-country Gini | 0.0080 | 0.0046 | -0.0035 |

Across the three sampled imputations, the within-country coefficients varied modestly relative to their ordinary model-based uncertainty:

| Model | SD across three imputations | Mean model SE |
|---|---:|---:|
| M3 within-country Gini | 0.0042 | 0.0096 |
| M4 within-country Gini | 0.0039 | 0.0094 |

The sample was identical across the three imputations and none of the six fitted models was singular or produced a convergence warning.

These results indicate no evidence that restricting the analysis to in-person interviews creates a qualitatively different sensitivity to SWIID measurement uncertainty.

The three-imputation results are treated strictly as a validation exercise. Their Rubin-pooled confidence intervals, p-values, degrees of freedom and fractions of missing information are not treated as final inferential quantities.

A full 100-imputation in-person analysis was not repeated because:

1. the complete 100-imputation Rounds 1–11 analysis had already established that SWIID uncertainty does not materially alter inference;
2. the in-person analysis addresses a separate survey-mode concern;
3. the three-imputation validation showed stable samples, clean model fitting and close agreement with the summary-series in-person estimates.

Repeating all 100 imputations would therefore amount primarily to crossing two already evaluated sensitivity dimensions rather than addressing a distinct unresolved source of uncertainty.

## 19. Reporting decision

The preferred reporting strategy is therefore:

- use the SWIID summary-series models for the main presentation;
- report the full 100-imputation Rounds 1–11 analysis as the principal sensitivity analysis for SWIID measurement uncertainty;
- retain the full 100-imputation Rounds 1–9 results as a historical-sample robustness check;
- report the summary-series Rounds 1–11 in-person model as the interview-mode sensitivity;
- describe the three-imputation in-person exercise as a validation showing no indication that SWIID uncertainty behaves anomalously under the mode restriction.

The results support use of the SWIID summary series for the primary model presentation because explicit uncertainty propagation does not materially change the estimated inequality coefficients, their precision or the substantive conclusions.

This does not imply that SWIID measurement uncertainty is negligible.

For the within-country Gini coefficients, approximately 37–41% of the Rubin total variance is attributable to between-imputation variation in the two full analyses.

Rather, the substantive conclusion is that this additional source of uncertainty is not large enough to alter inference relative to the summary-series specifications.

## 20. Reproducibility and interpretation

The executable scripts remain authoritative.

The uncertainty audit verifies the SWIID inputs before modelling.

The Rounds 1–9 and Rounds 1–11 model scripts reconstruct contextual exposures separately within each SWIID imputation, recalculate the within-between decomposition and pool the fitted fixed effects using Rubin's rules.

The in-person validation preserves the same contextual exposure definition as the main Rounds 1–11 analysis and restricts only the respondent-level estimation sample.

This notebook documents the methodological decisions connecting those stages and records the completed sensitivity results.

The SWIID uncertainty analysis addresses uncertainty in the inequality exposure.

It does not address:

- uncertainty in V-Dem political-discourse measures;
- uncertainty in World Bank macroeconomic controls;
- alternative survey-weighting estimands;
- broader model-specification uncertainty.

Those issues require separate sensitivity analyses.